Subsystem vs Multipole

### 2021-11-14 multipoleSSB/subsystem.tex
first pull from overleaf

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\renewcommand{\max}{\text{max}}
\newcommand{\goesto}{\rightarrow}

\title{Breaking monopole symmetry to dipole symmetry}
\author{charlie}
\date{October 2021}

@@

\section{Fields}

We have a field $\phi$ that transforms into $\phi + a^\mu x_\mu + b$ under a dipole symmetry. If $\phi$ is well-defined and constant, then the symmetry is fully broken. We only want to consider fields with constant values. How can we define a field that breaks the symmetry group to the monopole group? It must change under the $a^\mu$ part of the transformation, but somehow be rotationally invariant.

It is tempting to use a field like $\partial_\mu \phi$. The problem then is that if $\partial_\mu \phi$ is well-defined, then $\phi$ must be as well. Alternatively, it is tempting to use a vector-valued field $\theta_\mu$, but this will necessarily break rotation symmetry. Something like $\theta_\mu \theta^\mu$ would be rotationally invariant, but the ground state would still have to choose a value of $\theta_\mu$.

Maybe it would help to look at the group structure. The group $\mathcal{M}^2_\max$ has a space subgroup, a dipole subgroup $U(1)^d$, and a monopole subgroup $U(1)$. The space subgroup and the monopole subgroup combine to form $\mathcal{M}^1_\max$, but the dipole and space subgroups cannot combine without including the monopole subgroup. This should not be a problem though, as $\mathcal{M}^1_\max$ is a perfectly valid subgroup.

If we don't want to break rotational invariance, our order parameter needs to be a scalar\dots

\section{On the lattice}

Ignoring the question of whether we can do this in the continuum, can we partially break the symmetry on the lattice? The system
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ A  n_j + B n_j^2 \right] \nn
&\qquad + \sum_{j,\mu,\nu} C \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ A' n'_{j,j+\mu} + B' (n'_{j, j + \mu})^2 \right]\nn
&\qquad  + \sum_{j,\mu,\nu} C' \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
\emph{almost} manages to break dipole without breaking rotations. Say that in the dipole condensate we have $d_1=e^{i\phi_1}$, $d_2=e^{i\phi_2}$, and $d_3=e^{i\phi_3}$.

>>>>

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\goesto}{\rightarrow}

\title{Subsystem vs Multipole}
\author{charlie}
\date{October 2021}

@@

For subsystem symmetries we have the conservation law
\begin{align}
\frac{d}{dt} \int_{y=a} dx \; \rho = \frac{d}{dt} \int_{x=b} dy \; \rho = 0.
\end{align}
This is equivalent to the conservation law
\begin{align}
0 &= \frac{d}{dt} \int dxdy \, \big ( f(x) + g(y) \big )\rho,
\end{align}
for arbitrary functions $f$ and $g$. We can then Taylor expand these functions. This makes the conservation law look like a conservation law for a multipole symmetry where the polynomials are $x, x^2, x^3, \dots$ and $y, y^2, y^3, \dots$
